Exploring probability distributions for bivariate temporal granularities

### 2019-08-05 main.Rmd
multiple from single

The first condition implies that the granules in a linear granularity are non-overlapping and their index order is same as time order. \autoref{fig:linear-time} shows the implication of this condition.. If we consider the bottom linear granularity [@aigner2011visualization] as hourly and the entire horizon has T hours then it will have $\lfloor T/24\rfloor$ days, $\lfloor s/(24*7)\rfloor$ weeks and so on.

@@

Suppose we have a tsibble [@Wang2019-aw] with a time index in one column and keys and variables in other columns.  A time domain, as defined by [@Bettini1998-ed], is essentially a mapping of row numbers (the index set) to the time index.  A linear granularity is a mapping of row numbers to subsets of the time domain. For example, if the time index is days, then a linear granularity might be weeks, months or years.

For circular granularities, we need a symbolic representation of time to represent periodicity, which we call 'calendar categorization'. Anything that maps a time index to a categorical variable can be considered a calendar categorization. The number of categories is essentially the periodicity of a circular time granularity.

We want to use modular arithmetic on the domain of the circular granularity to define the calendar categorization. Hence, we start with the definition of equivalence classes and then move on to define a circular granularity.

@@

**Definition:** A **circular granularity** $C$ with a modular period m is defined to be a mapping from the integers $Z$ (Index Set) to $Z_m$, such that $C(s) = (s\mod m)$ for $s \in Z$.

@@

A few categorizations are listed below using modular arithmetic with appropriate period to illustrate further. We assume that the time index is in hours, and $n_i$ is the number of categories created by $C_i$. Then the following categorizations can be computed.

@@

## Computation of mnultiple order-up granularities from single order-up granularities

@@

Then, 
$f(x, y) & = f(x, x+1) + c(x, x+1)*[f(x+1, y)-1] $, for for $order(x+1) < order(y)$ can be used recursively to obtain any multiple order-up granularities. For $order(x+1) =  order(y)$, the function is same as the single-order-up granularities.

\begin{equation} \label{eq1}
\begin{split}
f(x, y) & = f(x, x+1) + c(x, x+1)*[f(x+1, y)-1] \\  
  & =  f(x, x+1) + c(x, x+1)[f(x+1, x+2) + c(x+1, x+2)[f(x+2, y) - 1] - 1]\\ (expanding f(x+1, y))  
  & =  f(x, x+1) + c(x, x+1)[f(x+1, x+2) - 1] + c(x, x+1)c(x+1, x+2)[f(x+2, y) - 1]
\end{split}
\end{equation}

@@

  $f(w,y) = f(x,y) \mod c(x,w)$, for $f(x,y) \mod c(x,w)!=0$ \\
  $= c(x,w))$, otherwise.

>>>>

The first condition implies that the granules in a linear granularity are non-overlapping and their index order is same as time order. \autoref{fig:linear-time} shows the implication of this condition. If we consider the bottom linear granularity [@aigner2011visualization] as hourly and the entire horizon has T hours then it will have $\lfloor T/24\rfloor$ days, $\lfloor s/(24*7)\rfloor$ weeks and so on.

@@

Suppose we have a tsibble with a time index in one column and keys and variables in other columns. A time domain, as defined by Bettini, is essentially a mapping of row numbers (the index set) to the time index. A linear granularity is a mapping of row numbers to subsets of the time domain. For example, if the time index is days, then a linear granularity might be weeks, months or years.

What we need to add to this are additional categorizations of time that are not linear granularities and are useful to represent periodicity. Examples include day-of-week, time-of-day, week-of-year, day-of-month, month-of-year, workingday/non-workingday, etc. Many of these are circular, such as day-of-week, time-of-day. Some are nearly circular such as day-of-month. Some are irregular such as workingday/non-working day. Let's call all of these "calendar categorizations". Anything that maps a time index to a categorical variable can be considered a calendar categorization. We can specify the categorizations corresponding to circular granularities using modular arithmetic. The number of categories is essentially the periodicity of a circular time granularity.

For example, suppose the time index is in minutes, and let $n_i$ be the number of categories created by $C_i$. Then the following categorizations can be computed.

\begin{table}[ht]
\begin{center}
\begin{tabular}{lr@{~}lr@{~}r}
\toprule
MOH: & $C_1(s)$ & $= s \mod 60$ & $n_1 =$&$60$ \\
MOD: & $C_2(s)$ & $= s \mod `r 60*24`$ & $n_2=$&$`r 60*24`$\\
MOW: & $C_3(s)$ & $= s \mod `r 60*24*7`$ & $n_2=$&$`r 60*24*7`$\\
HOD: & $C_4(s)$ & $= \lfloor s/60\rfloor\mod 24$  & $n_4 =$&$24$ \\
HOW: & $C_6(s)$ & $= \lfloor s/60\rfloor\mod 24*7$  & $n_5=$&$`r 24*7`$\\
DOW: & $C_7(s)$ & $= \lfloor s/24*60\rfloor \mod 7$ & $n_5=$&$7$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Illustrative calendar categorizations with hourly time index}
\label{definitions}
\end{table}

We start with the definition of an equivalence class and then move on to define a circular granularity.

@@

**Definition:** A **periodic circular granularity** $C$ with a modular period m is defined to be a mapping from the integers $Z$ (Index Set) to $Z_m$, such that $C(s) = (s\mod m)$ for $s \in Z$.

@@

**Definition:** A **aperiodic circular granularity** can not be defined in terms of modular arithmetic since each coarser unit is not always made up of equal number of finer units.

@@

## Computation of multiple order-up granularities from single order-up granularities

@@

Then, 
$f(x, y) = f(x, x+1) + c(x, x+1)[f(x+1, y)-1]$, for $order(x+1) < order(y)$ can be used recursively to obtain any multiple order-up granularities. For $order(x+1) =  order(y)$, the function is same as the single-order-up granularities.

\begin{equation} \label{eq1}
\begin{split}
f(x, y) & = f(x, x+1) + c(x, x+1)(f(x+1, y)-1) \\  
  & =  f(x, x+1) + c(x, x+1)[f(x+1, x+2) + c(x+1, x+2)(f(x+2, y) - 1) - 1]\\
  & =  f(x, x+1) + c(x, x+1)(f(x+1, x+2) - 1) + c(x, x+1)c(x+1, x+2)(f(x+2, y) - 1)\\
  & =  f(x, x+1) + c(x, x+1)(f(x+1, x+2) - 1) + c(x, x+2)(f(x+2, y) - 1)\\
  &\vdots\\
  & = \sum_{i=0}^{order(y) - order(x) + 1} c(x, x+i)(f(x
  +i, x+i+1) - 1)\\
  & = \sum_{i=0}^{order(y) - order(x) + 1} c(x, x+i)(f_1(x+i) - 1)
\end{split}
\end{equation}

This is valid in a scenario when the relationship between x and (x+1) is periodic (Define periodic in Definition section). However, when x and x+1 is not periodic, then some modification is required in Equation(1).

Consider, a hierarchy table of the form below:

```{r, hierarchy_tbl}
hierarchy_model <- tibble::tibble(units = c("minute", "hour", "day", "month", "quarter", "semester", "year"), conversion_fac  = c(60, 24, NA, 3, 2, 2, 1))
knitr::kable(hierarchy_model)
```

The table suggests that one hour is composed of 60 minutes, 1 day is composed of 24 hours, 1 quarter is composed of 3 months, 1 semester is consists of 2 quarters and 1 year is made up of 2 semesters. These partitions are periodic, since the number of finer units within each of these coarses units are same always.

@@

  $f(w,y) = f(x,y) \mod c(x,w)$, for $f(x,y) \mod c(x,w)!=0$
  $= c(x,w))$, otherwise.
